Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-10-14 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
ModelBasedTestCaseGenerationDataStreamProtocol paper: revisione

  Sensor equipped devices typically communicate with a streaming
  packet-based network protocol.  The communication is focused on
  controlling the sensors and report the sensors measurements.  The
  \STSB is an example of such a device, it communicates with an
  arbitrary computer system, the \STPU over a USB bus using a
  streaming packet-based protocol.  The \STU is a custom-designed, but
  inexpensive, sensor platform: it is composed by a one or more \STSBs
  paired with a general-purpose small-scale compute node, the \STPU
  (including everything from PDAs to powerful touchscreen-based
  portable computers).  There are over a dozen sensors on the \STSB
  itself, and new sensors can be added either using expansion slots on
  the \STSB, or adding them directly to the USB bus.  A third-party
  low-level library, provided by the USB chip vendor, the USB chip
  that has been used in the \STSB, creates and parses generic USB
  communication packets.

  A custom medium-level Java driver creates and parses the \STSB
  streaming communication packets.  The driver is formally specified,
  validated and verified with rigorous unit testing, both hand made
  and automatically generated.  Because of this process and
  combination of technologies, using this specification we are able to
  both verify a real \ST and a simulated one.  The system have been
  developed concurrently with the protocol specification and the real
  \STSB, the board were not available until the last months of
  development.  The protocol model is itself an executable stream
  simulator that can be used to develop new \ST user applications
  without having the \ST hardware at all.  Consequently, from this
  single specification, we can test that the \ST hardware communicates
  as expected, test that our \ST simulator communicates as expected,
  and simulate the \ST protocol stream itself.

@@

This system integrates a sensor platform, a datastore, and a compute
farm.  The sensor platform is a custom-designed, but inexpensive,
sensor platform (called the \ST) paired with general-purpose
small-scale compute nodes, including everything from low-power PDAs to
powerful touchscreen-based portable computers.  There are over a dozen
sensors on the \ST itself, and new sensors can be added by adding them
to the board.

Because of hard time constraints, and the initial unavailability of
the custom board, we needed to progress concurrently with all the
development tasks: the specification of the communication protocol
with the \STSB, the development of the controlling code in the \STSB,
the development of the driver to communicate with the \STSB, and the
development of the \STSB simulators needed for future tests of the
system (the board itself with the embedded software controlling it was
developed by an external manufacturer, hence, it is not directly taken
into account here).

There exist various approaches in literature addressing this kind of
development constraints (rapid development in hardware software
co-design), but they are all focused on big scale systems.  The
hardware software formal co-design methodologies usually have a common
structure\cite{Slomka2000,Hoffman2001}:

@@

Even when considering the development of the device hardware, embedded
software and software driver as separate entities that possibly have
to be developed concurrently, the approaches proposed are similar in
structure to the one seen in the case of hardware software
co-design\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}: they all focus
on very accurate formal specification that drive all the other
development tasks.

This approach is not feasible or convenient in our case: we want to
develop all the software components concurrently, we want to enhance
the protocol specification incrementally; we are not able to provide a
complete specification from the beginning, we are not interested in
direct code generation from specification, we want to generate
simulators to be used to feed the remaining system, even if the \STSB
is not available, and we want to be able to verify the \STSB behavior
when it is available.

To tackle the complexity of the concurrent tasks the formal model of
the protocol was used as a focus point that led all the other
development tasks (in this part, the approach is similar to the one
used in rapid development in hardware software co-design, as mentioned
above).  The protocol to specify has been divided in various (thin)
layers, each of the layer has been specified with a different
(lightweight) approach, some of the layers are completely formalized.
The formal model of the protocol was used to check at run time the
correctness of the /SB, the driver, the simulator, and the hand made
test suites of the driver and the simulator.  The formal model was
used to automatically generate a test suite to verify that both the
driver and the simulator were compliant to the specification, and
eventually to verify the specification itself.

The approach proved itself effective, at the point that, when the
board was finally released, only one small defect were found in the
driver or the simulator, while the errors and not agreed changes
introduced in the board where discovered with no effort.  Besides, the
simulators were able to feed with correct data the rest of the system

@@

In section \ref{sec:data_stream_and_protocol_description} the data
stream model and protocol is presented.

In section \ref{sec:data_stream_and_protocol_specification} the data
stream model and protocol is specified.

In section \ref{sec:test_cases} the test cases structure is presented,
for both the hand made and the generated ones.  Both the approaches
are analyzed, their pro and contra are discussed, as well as the
expected advantages to use them in combination.

In section \ref{sec:test_cases_on_the_job} the execution of the test
suites is described, the results are presented.

In section \ref{sec:test_cases_retrospectives} the data collected and
the results are analyzed, both from a qualitative and a quantitative
points of view.

In section \ref{sec:related_works} related works are presented and
analyzed.

Finally in section \ref{sec:conclusion} conclusions are drawn and
ongoing and future works on similar topics is introduced.

@@

The communication between \STSB and \STPU is realized through an
Universal Serial Bus (USB) interface.  On top of the USB serial
channels a proprietary protocol is used, specified by
FTDI\footnote{\myhref{http://www.ftdichip.com/}{FTDI web site}}; FTDI
is the manufacturer of the USB controller chip used in the \STSB, the
FT2232H\cite{ftdi_ft232h_2009}.

The proprietary protocol enables the control and management of some of
the functionalities in FT232H chip (such as internal EEPROM read and
write, reset the chip, etc.) as well as the definition of the
communication protocol and properties to be used in the asynchronous
channels (input and output channels) provided by FT232H.  These added
functionalities are accessed via an Application Program Interface, the
API is provided by a driver developed directly by FTDI, which is
available for various systems (Linux 32 and 64 bits, Mac OS, Windows)
\cite{ftdi_d2xx_api_2009}.

Finally, on top of the serial channels provided by the API, a custom
protocol has been developed to manage the entire \STSB; that is, read
sensor data observations, enable or disable sensors, configure
on-board components as the Field-Programmable Gate Array (FPGA), and
the micro processor; the protocol main function is to read sensor
data.

The protocol is an asynchronous fixed length packed based protocol.
The protocol has been divided in multiple layers to ease its formal
description.  The rationale is the following: mixing up different
abstraction levels of a protocol description can lead to overly
complex and unmanageable descriptions, especially when dealing with
formal ones.  The layers identified in protocol are the following:

@@

We will only examine the output sensor data packet, since it is the
more used and interesting one.  We will examine the first three
layers, that is: the packet byte structure, the packet info structure
and the single packet rules.

The packet byte and info structures of the output sensor data packet
reflects the board capabilities and sensors.  Basically a single
packet has to accommodate various types of data:
\begin{itemize}
\item Fast data rate streams: audio data, 48KHz or 96KHz.
\item Medium data rate streams: 5KHz.
\item Slow data rate streams: temperature, pression, accelerometer,
  etc.; all these sensor have a data rate far lower than 5 KHz.
\item Metadatas: metadatas on sensors and board state.
\end{itemize}

Even if we have to deal with data streams that have different data
rates, the internal structure of the package (as we will see) is
strictly fixed.  This choice originates from a trade-off between two
constraints: the efficiency of the transmission (redundant data to be
minimized) and the complexity to build a package (complexity to be
minimized).  The complexity minimization constraint rationale is the
following: the FPGA on the \STSB builds the packet, but the FPGA is
also to be used for various other specific tasks; it is important to
maintain the building packet code on the FPGA as simple and short as
possible, to leave enough space to implement other tasks.  The Output
sensor data packet structure is showed in Table
\ref{tab:data_packet_structure}, the packet total length is 1024 bytes
(512 16 bit words).

\begin{table}[htbp]
  \caption{Output sensor data packet structure.}
  \label{tab:data_packet_structure}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      \textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
      0-1 & sentinel\\\hline
      2-7 & slow data\\\hline
      8-11 & metadata: board state\\\hline
      12-13 & reserved\\\hline
      14-15 & metadata: board state\\\hline
      16 & reserved\\\hline
      17-22 & frame 1\\\hline
      \ldots & \ldots\\\hline
      503-508 & subsequence 82\\\hline
    \end{tabular}
  \end{center}
\end{table}

The frames accommodate data from the fast and medium data rate streams
(there are 4 fast data rate stream and 8 medium data rate stream
channels).  There are 82 frames in a packet.  The frame structure is
shown in Table \ref{tab:subsequence_structure}.

\begin{table}[htbp]
  \caption{Output sensor data packet subsequence structure.}
  \label{tab:subsequence_structure}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      \textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
      0 & metadata: subsequence description\\\hline
      1 & medium data rate data sample\\\hline
      2-5 & fast data rate data samples\\\hline
    \end{tabular}
  \end{center}
\end{table}

Looking at the structure described and at the data rates frequencies,
it can be seen that there must be something subtle, to make all the
figures to add up, that is, to be capable to accommodate three
different rates outputs in a fixed structure, fixed length packet.
The subtleties lies in two forms of redundancies, for slow and medium
data rates:

\begin{itemize}
\item Slow data rates: slow data can be found at the beginning of each
  packet, the data is simply repeated (that is, it is equal to the
  previous packet in the stream) when no new data is available.
\item Medium data rates: the 8 medium rate channels are multiplexed
  together into the medium data rate data sample of each subsequence,
  which channel the sample belongs to is a piece of info to be
  extracted from the subsequence metadata. Besides, the medium data
  rate data sample could contain no data at all, this info again is to
  be extracted from the metadata.
\end{itemize}

The single packet rules delimit the boundaries of the values
obtainable from the packet: each defined sensor represented in the
packet, as well as the metadata describing the \STSB and the packet
and frame contents, have a defined range of acceptable values.  There
are also rules affecting more than one value, basically validity rules
(i.e. if a specific sensor is active than the reading value has to be
in a defined range), and rules specifying a correct sequence of frames
(i.e. there can be only one switch in sampling rate in the frame
sequence of a single packet).

@@

The specification of the protocol has been used as a focus point to
lead all the other development tasks.

We needed a specification capable of:

\begin{itemize}
\item Be executed at run time. If something on the protocol is not
  respected, the run time execution environment is capable to detect
  and signal the fault.
\item Generate tests.
\item Verify simulators.
\item Be used in conjunction of hand written unit tests.
\end{itemize}

We also needed a specification capable to tackle properly all the
protocol layers.  Since the different layers have very different
characteristics, different context dependent specification approaches
were chosen.

@@

The specification is provided by a reference Java implementation
capable of recognizing the proper packet structure in a byte data
stream. The implementation is tricky, since it checks for the
sentinels repeatedly to be sufficiently confidant that the byte stream
can be properly decomposed into a packet stream.

The implementation can be obviously executed at run time (the
specification is the implementation code).  Meaningful automatic tests
cannot be generated.  The implementation can be used to verify a
packet byte structure simulator.  The implementation were actually
driven (Test Driven Development) by the test suite.

@@

The specification is provided both by the Java API that has been
built, and by the annotation with JML language\cite{JML2}\footnote{
  \myhref{http://www.cs.ucf.edu/~leavens/JML/}{The Java Modeling
    Language (JML)}}.  JML has been chosen because it is capable to
specify Java code: JML is a behavioral interface specification
language.  JML specification language combines the design by contract
approach (applied to Java language) and the model-based specification
approach of interface specification languages.

Many tools are available supporting JML specification language,
focusing on different aspects of the language.  JML2 suite of
tools\cite{BurdyEtal05-STTT}
\footnote{\myhref{http://sourceforge.net/projects/jmlspecs/}{Common
    JML tools}} was chosen, since both run time verification support
and test generation support are provided directly by the suite.

The Java API specification part is checked automatically by the Java
compiler, while the JML specification part can be checked at run time
using the provided JML2 run time environment.  Meaningful automatic
tests can be generated from the JML specification.  The JML
specification can be used to verify an abstract packet generation
simulator, making the simulator an extension of the Java interfaces.
The specification can also be used to verify the real driver
implementation, based on the packet byte structure implementation
presented above.  The implementation was driven by the JML
specification and by an hand made test suite.  Finally, hand made unit
tests can be checked, during execution, against the specification,
using the provided JML2 run time environment.

@@

The specification is provided exclusively by annotations with JML
language.

The capabilities of a JML specification have been already described in
the previous paragraph, that is: specification checkable at run time,
tests generation, simulator verification, usable with hand written
tests.

@@

Simulators are needed to test in isolation parts of the system;
especially the upper layers, yet to be developed.  During the
development the simulators were essential, since the board was not
available.  Simulators were built concurrently with the specification,
the real driver code, and the tests.

Theoretically a separate simulator should be needed for each protocol
layer; a separate simulator for each layer would enable us to
separately specify, develop and test each layer of the protocol.  In
our case we evaluated not worth the effort to maintain such a
separation in the simulators, motivated by the fact that the protocol
layers identified are too fine grained to be separated into different
simulators.  Two simulators were built.

The two simulators were built reflecting the software abstraction
layers defined for the driver.  The driver API is splitted in two
structural layers: the general high level interfaces, that expose the
main functionalities of the board and the main contents of the packets
hiding the details, and the lower level implementation that parse the
data streams in and out the board, to translate the higher level
instructions and data in properly formed packets.  Hence, we have
implemented two simulators: one high level simulator, which
re-implements the high level interfaces providing corresponding high
level mechanisms to control the data and the behavior of the simulated
board, it has nothing to do with the details of the data streams.  One
low level simulator which is capable to rebuild the data stream out of
the board based on the data stream into the board, the in and out data
streams are built exactly as the sensor board is expected to be able
to parse or generate.

@@

The JML specifications used to annotate Java code are of varying
complexities.  Some of the specifications are simple, focusing on
constraints that should hold when calling a method (preconditions) and
the constraints on the return value (a very basic form of
postconditions).  In listing \ref{lst:jml_example_simple} a simple JML
specification example is shown, the method \lil{getTemperature} is
declared to be \lil{/*@ pure @*/}, which means that it cannot change
the state of any object (a postcondition), the specification also
constraints the return value with a lower and upper bound (another
postcondition).

@@

Some of the specifications are rather complex, usually focusing on
properties regarding the behavior of a whole object.  In listing
\ref{lst:jml_example_invariant} a complex invariant (a property that
has to be maintained, that is, that is assumed on entry and guaranteed
on exit of each method of an object) example is shown.  The invariant
is constraining the number of samples for each medium data rate
streams: the total number of streams is \lil{Frame.ADC_CHANNELS}, the
total number of frames is \lil{FRAMES}, the constant constraining the
number of samples is \lil{FRAMES/Frame.ADC_CHANNELS+1}, meaning that
the samples contained in a frame are fairly distributed on the
channels.  The valid samples are counted parsing all the frames
contained in a packet, selecting only the matching valid samples.  A
medium data rate stream sample is considered valid when
\lil{isADCActive()} method returns \lil{true}.

@@

The unit test cases that have been built to verify the protocol driver
are of two kinds: hand made unit tests and unit tests automatically
generated based on JML specifications.  The tests have been thought to
be used together, they are complementary.  The tests effectiveness has
been evaluated, for each kind of test, the evaluation is carried out
in section \ref{sec:test_cases_retrospectives}.  The test
effectiveness evaluation has been based on three elements: effective
results on piloting the real board (quantitative), code coverage
(quantitative), development help and usefulness (qualitative).

The dual testing approach was thought as a winning option, when
compared to hand based tests or completely automated tests used alone,
testing a driver for a stream communication protocol with hard
constraints on development time.  \ST stream communication protocol is
characterized by a well definite packet structure, and a well definite
semantic on the packet contents as well.  On the other hand in the
protocol there are not strong constraints on synchronization, since
the communication channel is splitted in two (in and out) separate
concurrent and asynchronous channels.  The lack of hard
synchronization constraints reduce the advantage to use a classical
state machine based specification (as the one presented in
\cite{Hubbers2004}).

Hard development time constraints were also a precise requirement; the
time constraints pushed us towards a incomplete (\emph{just good
  enough}) specification.  In \STSB driver JML was used to specify
very simple behaviors (method oriented) on the whole driver
implementation, and more complex behavior on the driver core
interfaces.  The software system parts not specified by the JML2
specification were specified using hand made unit tests.

Hand made unit tests cover behavior not covered by the automatically
generated tests, that is, they cover the not specified behavior; but
this is not the only purpose they have.  Hand made unit tests are used
to drive the development, since the Test Driven Development
approach\cite{beck2003test} has been followed during development.
Hand made tests are also used to drive the development of the most
error prone or complex JML2 specifications; a test developed to drive
the specification is a test that should fail if the specification is
not working as expected.

@@

The package structure of the driver developed is shown in Figure
\ref{fig:class_diagram_main}: two independent packages has been
defined: \lil{Stream} and \lil{Driver}.  The packages are abstract,
that is, they mainly contain abstractions; in Java language this is
translated into a package which contains mainly abstract classes or
interfaces.  The two abstract packages has various implementations,
the dependency between abstract packages is not direct, but only
through an implementation (\lil{StreamDriver}).  This is needed to
maintain a high decoupling of the packages, and is the result of
applying the \emph{classical} dependency inversion
principle\cite{martin1996dependency} (the dependency structure of the
packages has a saw shape, a common shape on systems where the
dependency inversion principle has been followed).

@@

The JML specification has been used to generate the tests.  A specific
tool of the JML2 suite can be used to automatically generate
them\cite{Cheon-Leavens02}.

>>>>

  Sensor equipped devices typically communicate with a streaming
  packet-based network protocol.  The communication is focused on
  controlling the sensors and reporting the sensors measurements.  The
  \STSB is an example of such a device: it communicates with an
  arbitrary computer system, the \STPU, over a USB bus using a
  streaming packet-based protocol.  The \STU is a custom-designed, but
  inexpensive, sensor platform: it is composed by a one or more \STSBs
  paired with a general-purpose small-scale compute node, the \STPU
  (including everything from PDAs to powerful touchscreen-based
  portable computers).  There are over a dozen sensors on the \STSB
  itself, and new sensors can be added either using expansion slots on
  the \STSB, or adding them directly to the USB bus.  A third-party
  low-level library creates and parses generic USB communication
  packets; the library has been provided by the vendor of the USB chip
  vendor, the USB chip that has been used in the \STSB.

  A custom medium-level Java driver creates and parses the \STSB
  streaming communication packets.  The driver is formally specified,
  validated and verified with rigorous unit testing, both hand made
  and automatically generated.  Because of this process and
  combination of technologies, we are able to verify a real \ST and a
  simulated one as well.  The system has been developed concurrently
  with the protocol specification and the real \STSB, the boards have
  not been available until the last months of development.  The
  protocol model is itself an executable stream simulator that can be
  used to develop new \ST user applications without having the \ST
  hardware at all.  Consequently, from this single specification, we
  are able to test that the \ST hardware communicates as expected,
  that our \ST simulator communicates as expected, and simulate the
  \ST protocol stream itself.

@@

This system integrates a sensor platform, a datastore, and a compute
farm.  The sensor platform is a custom-designed, but inexpensive,
sensor platform (called the \ST) paired with general-purpose
small-scale compute nodes, including everything from low-power PDAs to
powerful touchscreen-based portable computers.  There are over a dozen
sensors packaged on the \ST itself, the board is expandable as well,
new sensors can be added to it easily.

Because of hard time constraints, and the initial unavailability of
the custom board, we needed to progress concurrently with all the
development tasks:

\begin{itemize}
\item specification of the communication protocol;
\item development of the physical board specification and following
  realization;
\item development of the board embedded software;
\item development of the software driver to communicate with the
  board;
\item development of the software simulators of the board.
\end{itemize}

The development of the physical board, along with its embedded
software, has been done by an external manufacturer, thus it is not
directly taken into account here.

There exist various approaches in literature addressing this kind of
development constraints (rapid development in hardware software
co-design), but they are all focused on big-scale systems.  The
hardware software formal co-design methodologies usually have a common
structure\cite{Slomka2000,Hoffman2001}:

@@

Even when considering the development of the device hardware, embedded
software and software driver as separate entities, that possibly have
to be developed concurrently, the approaches proposed are similar in
structure to the one seen in the case of hardware software
co-design\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}: they all focus
on a specification that aims to be as complete as possible. The
specification, once it is finalized, is used to define the remaining
development tasks, through automatic code generation, refinement,
model checking, etc.

This approach has been considered not feasible nor convenient if
applied to our case since it does not respect some of the
requirements: first of all, we want to develop all the software
components concurrently, we want to enhance the protocol specification
incrementally and we are not able to provide a complete specification
from the beginning, hence, a properly defined specification is not
available in advance. Secondly, we are not interested in direct code
generation from specification, since we do not want to have any
dependencies on code generators; in fact the programming language to
be used to develop the driver has been postponed for a while.

To tackle the complexity of the concurrent development tasks, the
formal model of the protocol was used as a focus point that led all
the other development tasks. As it can be seen, the central role of
the specification is unchanged; what is changed is how and when the
specification is developed.

The protocol to specify has been divided in various (thin) layers,
each of the layer has been specified with a different lightweight
approach: some of the layers have been specified formally, while
others have been specified in an unprecise way.  The formal
specifications have been used to check at run time the correctness of
the \SB, the driver and the simulators.  The formal specifications
have also been used to automatically generate a test suite to verify
the driver and the simulators, and eventually to verify the
specification itself.  An hand made test suite has been used to verify
the driver and the simulators as well.  The generated and hand made
test suites have been used together, complementing each other.

The approach proved itself effective, at the point that, when the
board has been finally released, only one small defect has been found
in the driver, while the errors and not agreed changes introduced in
the board and in the protocol it implemented, have been discovered
immediately, with no additional effort.  Besides, the simulators are
able to feed with correct data the rest of the system.

@@

In section \ref{sec:data_stream_and_protocol_description} the data
stream model and protocol are presented.

In section \ref{sec:data_stream_and_protocol_specification} the data
stream model and protocol are specified.

In section \ref{sec:test_cases} the test cases structure is presented,
for both the hand made and the generated test suites.  Both the
approaches are analyzed, their pro and contra are discussed, as well
as the expected advantages to use them in combination.

In section \ref{sec:test_cases_on_the_job} the execution of the test
suites is described; the results and measures obtained by their
execution are presented.

In section \ref{sec:test_cases_retrospectives} the data collected in
the previous section are analyzed, both from a qualitative and a
quantitative points of view. The effectiveness of the approach is
discussed as well.

In section \ref{sec:related_works} the related works that have not
been mentioned in the current section are presented and analyzed.

Finally in section \ref{sec:conclusion} conclusions are drawn, and
ongoing and future works on similar topics are introduced.

@@

\STSB and \STPU communicates via an Universal Serial Bus (USB)
interface.  A proprietary protocol is used on top of the USB serial
channels, the protocol is specified by
FTDI\footnote{\myhref{http://www.ftdichip.com/}{FTDI web site}}; FTDI
is the manufacturer of the USB controller chip used in the \STSB, the
FT2232H\cite{ftdi_ft232h_2009}.

The proprietary protocol enables the control and management of some of
the functionalities in FT232H chip (such as internal EEPROM read and
write, reset of the chip, etc.), as well as the definition of the
communication protocol and properties to be used in the asynchronous
channels (input and output channels) provided by FT232H.  These
functionalities are accessed via an Application Program Interface, the
API is provided by a driver developed directly by FTDI, which is
available for various systems (Linux 32 and 64 bits, Mac OS, Windows)
\cite{ftdi_d2xx_api_2009}.

Finally, a custom protocol has been developed to manage the entire
\STSB, built on top of the serial channels provided by the API.  The
custom protocol is meant to be used to read sensor data observations,
enable or disable sensors, configure on-board components (as the
Field-Programmable Gate Array and the micro processor); the protocol
main function is to read sensor data.

The protocol is asynchronous and packet based, the packets have a
fixed length.  The protocol has been divided in multiple layers to
ease its formal description.  The rationale is the following: mixing
up different abstraction levels of a protocol description can lead to
overly complex and unmanageable descriptions, especially when dealing
with formal ones.  The layers identified in protocol are the
following:

@@

We will examine the output sensor data packet, since it is the most
used and interesting one; we won't describe input or other output
packets.  We will examine the first three layers, that is: the packet
byte structure, the packet info structure and the single packet rules.

The packet byte and info structures of the output sensor data packet
reflect the board capabilities and built in sensors.  Essentially a
single packet has to accommodate various types of data:

\begin{itemize}
\item Fast data rate streams: audio data, 48KHz (standard audio) or
  96KHz (ultra audio).
\item Medium data rate streams: 5KHz.
\item Slow data rate streams: temperature, pression, accelerometer,
  etc.; all these sensor have a data rate lower than 5 KHz.
\item Metadatas: metadatas on sensors and board state.
\end{itemize}

The internal structure of the package (as we will see) is strictly
fixed, even if the data streams have different data rates.  This
choice originates from a trade-off between two constraints: the
efficiency of the transmission (redundant data to be minimized) and
the complexity to build a package (complexity to be minimized).  The
complexity minimization constraint rationale is the following: the
FPGA on the \STSB builds the packet, but the FPGA is also to be used
for various other specific tasks; it is important to maintain the
building packet code on the FPGA as simple and short as possible, to
leave enough space to implement other tasks.  The Output sensor data
packet structure is shown in Table \ref{tab:data_packet_structure},
the packet total length is 1024 bytes (512 16 bit words).

\begin{table}[htbp]
  \caption{Output sensor data packet structure.}
  \label{tab:data_packet_structure}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      \textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
      0-1 & sentinel\\\hline
      2-7 & slow data\\\hline
      8-11 & metadata: board state\\\hline
      12-13 & reserved\\\hline
      14-15 & metadata: board state\\\hline
      16 & reserved\\\hline
      17-22 & frame 1\\\hline
      \ldots & \ldots\\\hline
      503-508 & frame 82\\\hline
    \end{tabular}
  \end{center}
\end{table}

The frames accommodate data from the fast and medium data rate streams
(there are 4 fast data rate stream and 8 medium data rate stream
channels).  There are 82 frames in a packet.  The frame structure is
shown in Table \ref{tab:frame_structure}.

\begin{table}[htbp]
  \caption{Output sensor data packet frame structure.}
  \label{tab:frame_structure}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      \textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
      0 & metadata: frame description\\\hline
      1 & medium data rate data sample\\\hline
      2-5 & fast data rate data samples\\\hline
    \end{tabular}
  \end{center}
\end{table}

Looking at the structure described and the data rates frequencies, it
can be seen that there must be some subtle rule, to make all the
figures to add up; that is, to be capable to accommodate three
different rates outputs in a fixed structure, fixed length packet.
The subtleties lies in two forms of redundancies, for slow and medium
data rates:

\begin{itemize}
\item Slow data rates: slow data can be found at the beginning of each
  packet, the data is simply repeated (that is, it is equal to the
  previous packet in the stream) when no new data is available.
\item Medium data rates: the 8 medium rate channels are multiplexed
  together into the medium data rate data sample of each frame.  The
  medium data rate sample channel is a piece of info to be extracted
  from the frame description. Besides, the medium data rate data
  sample could contain no data at all, this information is to be
  extracted from the frame description. So, one frame could contain a
  medium data rate sample from channel 0, the following frame a sample
  from channel 5, the following frame no sample at all, etc.
\end{itemize}

The single packet rules delimit the boundaries of the values
obtainable from the packet: each defined sensor represented in the
packet, as well as the metadata describing the \STSB and the packet
and frame contents, are constrained by a range of acceptable values.
There are also rules affecting more than one value, these are
essentially validity rules (i.e. if a specific sensor is active the
reading value has to be in a defined range), and rules specifying a
correct sequence of frames (i.e. there can be only one switch in
sampling rate in the frame sequence of a single packet).

@@

The specifications of the protocol has been used as a central point to
lead all the other development tasks.

Regarding the formal specification used, we needed a specification
capable of:

\begin{itemize}
\item Be executed at run time: if something on the protocol is not
  respected, the run time execution environment is capable to detect
  and signal the fault.
\item Generate tests.
\item Verify simulators.
\item Be used in conjunction of hand written unit tests.
\end{itemize}

We needed specifications capable to tackle properly all the protocol
layers.  Since different layers have different characteristics,
different specification approaches has been chosen for each layer: the
specification are context dependent.

@@

The specification is provided by a hand made test suite, composed of
unit and integration tests; the tests specifies the behaviors of the
Java implementation, which is capable of recognizing the proper packet
structure in a binary data stream. The implementation is tricky: it
checks for the sentinels repeatedly to be sufficiently confidant that
the byte stream can be properly decomposed into a packet stream.

Fulfilled requirements:

\begin{itemize}
\item The specification cannot be checked at run time.
\item Meaningful automatic tests cannot be generated.
\item The specification can be used to verify a packet byte structure
  simulator.
\item The implementation were actually driven (Test Driven
  Development) by the test suite specification.
\end{itemize}

@@

The specification is provided by JML annotations\cite{JML2}\footnote{
  \myhref{http://www.cs.ucf.edu/~leavens/JML/}{The Java Modeling
    Language (JML)}} and by a hand made test suite.

JML has been chosen because it is capable to specify Java code: JML is
a behavioral interface specification language.  JML specification
language combines the design by contract approach (applied to Java
language) and the model-based specification approach of interface
specification languages.

Many tools able to support JML specification language are available,
focusing on different aspects of the language.  JML2 suite of
tools\cite{BurdyEtal05-STTT}
\footnote{\myhref{http://sourceforge.net/projects/jmlspecs/}{Common
    JML tools}} has been chosen, since the suite provides both run
time verification support and test generation support.

Fulfilled requirements:

\begin{itemize}
\item The Java API specification part is checked automatically by the
  Java compiler, while the JML specification part can be checked at
  run time using the provided JML2 run time environment.
\item Meaningful automatic tests can be generated from the JML
  specification.
\item The JML specification can be used to verify a simulator, which
  generates packets as the real board would do .The specification can
  also be used to verify the real driver implementation, based on the
  packet byte structure implementation presented above.
\item The implementation was driven by the formal JML specification
  and by the hand made test suite.  Finally, hand made unit tests can
  be checked, during execution, against the formal specification,
  using the provided JML2 run time environment.
\end{itemize}

@@

The specification is provided exclusively by the formal specification
with JML language.

The fulfilled requirements of a JML specification have been already
described in the previous paragraph, that is:

\begin{itemize}
\item The JML specification is checkable at run time.
\item Tests can be automatically generated.
\item Simulators can be verified.
\item The specification can be used with hand written tests.
\end{itemize}

@@

Simulators can be used to test parts of the system, to minimizing
their dependencies: a simulator can be used by upper layers, with no
need to provide the functionalities of the lower layers.  During the
development of the \STSB the simulators have been essential, since the
physical board was not available: we needed at least to remove the
dependencies over the real board to execute the driver implementation.
Simulators have been built concurrently with the specification, the
driver implementation, and the tests.

Theoretically a separate simulator should be needed for each protocol
layer: this would enable us to specify, develop and test each layer in
isolation.  In our case we evaluated not worth the effort to replicate
the layer structure into simulators; the choice was motivated by the
fact that such a distinction would have not provided us any
perceivable benefit.  Two simulators were built.

The simulators have been built reflecting the software abstraction
layers defined for the driver.  The driver API is splitted in two
abstraction layers: the general high level interfaces, that expose the
main functionalities of the board and the main contents of the
packets, and the lower level implementation that parse the data
streams in and out the board, meant to translate the higher level
instructions and data in properly formed packets.

The high level simulator implements the interfaces providing the
methods to deal with an abstracted \SB: the simulator knows nothing
about the details of the real format of the binary data streams.  The
low level simulator is capable to rebuild the \SB data streams: the in
and out data streams are built exactly as the sensor board is expected
to be able to parse or generate.

@@

The JML specifications that have been written are of varying
complexities.  Some of the specifications are simple, focusing on
constraints that should hold when calling a method (the preconditions)
and constraints on the return value (the very basic form of
postconditions).  In listing \ref{lst:jml_example_simple} a simple JML
specification example is shown, the method \lil{getTemperature} is
declared to be \lil{/*@ pure @*/}, which means that it cannot change
the state of any object (a postcondition), the specification also
constraints the return value with a lower and upper bound (another
postcondition).

@@

Some of the specifications are rather complex, usually focusing on
properties regarding the behavior of a whole object.  In listing
\ref{lst:jml_example_invariant} a complex invariant example is shown;
an invariant is a property that has to be maintained during the life
cycle of an object, more precisly, an invariant is assumed on entry
and guaranteed on exit of each method of an object.  The invariant is
constraining the number of samples for each medium data rate streams:
the total number of streams is \lil{Frame.ADC_CHANNELS}, the total
number of frames is \lil{FRAMES}, the constant constraining the number
of samples is \lil{FRAMES/Frame.ADC_CHANNELS+1}, meaning that the
samples contained in a frame are fairly distributed on the channels.
The valid samples are counted parsing all the frames contained in a
packet, selecting only the matching valid samples.  A medium data rate
stream sample is considered valid when \lil{isADCActive()} method
returns \lil{true}.

@@

The unit test cases that have been built to verify the protocol driver
are of two kinds: hand made unit tests and automatically generated
unit tests based on JML specifications.  The tests have been thought
to be used together, they are complementary.  The tests effectiveness
has been evaluated for each test suite; the evaluation is carried out
in section \ref{sec:test_cases_retrospectives}.  The test
effectiveness evaluation has been based on three elements: effective
results on piloting the real board (quantitative), code coverage
(quantitative), development help and usefulness (qualitative).

The dual testing approach was thought a winning option, when compared
to hand based tests or completely automated tests used alone.  \ST
stream communication protocol is characterized by a well definite
packet structure, and a well definite semantic on the packet contents
as well.  On the other hand, in the protocol, there are not strong
constraints on synchronization, since the communication channel is
splitted in two (in and out) separate concurrent and asynchronous
channels.  The lack of hard synchronization constraints reduces the
advantage to use a classical state machine based specification (as the
one presented in \cite{Hubbers2004}), hence this option has been ruled
out.

Hard development time constraints were also a precise requirement; the
time constraints pushed us towards an incomplete (\emph{just good
  enough}) specification.  In \STSB driver JML was used to specify
very simple behaviors (method oriented) on the whole driver
implementation, and more complex behavior on the driver core
interfaces.  The software system parts not specified by the JML2
specification were specified using hand made unit tests.

Hand made unit tests specifies the behavior not specified by formal
specification; nevertheless this is not the only purpose they have.
Hand made unit tests have been used to drive the development, since
the Test Driven Development approach\cite{beck2003test} has been
followed during development.

@@

The package structure of the driver developed is shown in Figure
\ref{fig:class_diagram_main}: two independent packages has been
defined (\lil{Stream} and \lil{Driver}).  The packages are abstract,
that is, they mainly contain abstractions; in Java language this is
translated into a package which contains mainly abstract classes or
interfaces.  The two abstract packages have various implementations,
and the dependency between the two is not direct, since it is realized
through an implementation (\lil{StreamDriver}).  This is needed to
maintain a high decoupling of the packages, and is the result of
applying the classical dependency inversion
principle\cite{martin1996dependency}: the dependency structure of the
packages has a saw shape, a common shape on systems where the
dependency inversion principle has been followed.

@@

The JML specification has been used to generate the tests.  A specific
tool of the JML2 suite is able to automatically generate
them\cite{Cheon-Leavens02}.

The test suite generated is a collection of fine grained unit tests.
For each public method a set of tests is generated.  The specified
behavior of the method is checked using the JML2 run time assertion
checker, hence, an oracle is not needed.  The generated tests send
messages to the method under test, once the initial precondition check is passed (if it is not passed, the test is discarded), the test catch assertion violation
exceptions;
if a precondition is not met the test is not run, if the
precondition is met the test is run and if the postcondition is not
met an exception is fired and the test fails.  The test data
used to build the tests is the only part that is required to be
provided by the user.
